---
title: "Downloading and caching BODS GTFS data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading and caching BODS GTFS data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bodsr2)
```

## Introduction

The [Bus Open Data Service (BODS)](https://data.bus-data.dft.gov.uk/)
provides timetable data for bus services in England as regional GTFS zip
files. These files are large — the East Midlands zip is around 600MB —
and downloading one on every session would be impractical.

`bodsr2` solves this with a simple caching layer. The first call to
`gtfs_load()` downloads the zip, parses it into a
[tidytransit](https://r-transit.github.io/tidytransit/) object, and
saves it to a local cache. Subsequent calls load directly from the
cache, which is dramatically faster.

## Loading GTFS data

```{r load, eval = FALSE}
library(bodsr2)

gtfs <- gtfs_load("east_midlands")
#> ℹ Downloading GTFS data for "east_midlands" from BODS.
#> ℹ This file is large and may take a moment.
#> ✔ Zip saved to ~/.cache/R/bodsr2/east_midlands.zip
#> ℹ Reading GTFS zip for "east_midlands". This may take a moment.
#> ✔ GTFS data cached to ~/.cache/R/bodsr2/east_midlands.rds
```

On a typical connection this takes around 40 seconds. The second call is
much faster:

```{r load-cached, eval = FALSE}
gtfs <- gtfs_load("east_midlands")
# completes in around 3 seconds
```

The cache is stored in your platform's user cache directory. You can
find it with `gtfs_cache_dir()`, and clear it with
`gtfs_delete_cache("east_midlands")`.

The cache is automatically refreshed when the GTFS feed expires
(indicated by `feed_end_date` in the feed info). You can also force a
fresh download at any time:

```{r force-download, eval = FALSE}
gtfs <- gtfs_load("east_midlands", download = TRUE)
```

## What's in the object

`gtfs_load()` returns a `tidygtfs` object — a named list of tibbles, one
for each GTFS table:

```{r names, eval = FALSE}
names(gtfs)
#>  [1] "agency"         "stops"          "routes"         "calendar"      
#>  [5] "calendar_dates" "trips"          "shapes"         "feed_info"     
#>  [9] "stop_times"
```

This is a standard tidytransit object. See the [tidytransit
documentation](https://r-transit.github.io/tidytransit/) for a full
guide to working with it. The rest of this vignette shows one example to
illustrate the kind of analysis the object supports.

## Example: services departing from Derby Bus Station Bay 4

The V1 Villager service (Derby to Burton-upon-Trent) departs from Bay 4
at Derby Bus Station. Its stop ID in the GTFS data is `"1090BSTN04"`:

```{r derby-stop, eval = FALSE}
gtfs$stops |>
  dplyr::filter(stop_id == "1090BSTN04") |>
  dplyr::select(stop_id, stop_name, platform_code)
#> # A tibble: 1 × 3
#>   stop_id    stop_name   platform_code
#>   <chr>      <chr>       <chr>        
#> 1 1090BSTN04 Bus Station 4
```

To find all services that start from this stop, we look for trips where
this stop appears at the lowest stop sequence number:

```{r bay4-services, eval = FALSE}
library(dplyr)

gtfs$stop_times |>
  filter(stop_id == "1090BSTN04") |>
  left_join(gtfs$trips, by = "trip_id") |>
  left_join(gtfs$routes, by = "route_id") |>
  distinct(route_short_name, trip_headsign)
#> # A tibble: 9 × 2
#>   route_short_name trip_headsign            
#>   <chr>            <chr>                    
#> 1 V1               Derby Bus Station        
#> 2 V1               Burton                   
#> 3 V1               Burton Train Stn         
#> 4 TP               Derby Bus Station (Bay 4)
#> 5 70               Derby Bus Station        
#> 6 70A              Derby Bus Station        
#> 7 70A              Methodist Church         
#> 8 V1               Wetmore                  
#> 9 70A              Willington
```

Four routes use Bay 4: the V1, TP, 70, and 70A. The `trip_headsign`
column shows the destination as displayed on the front of the bus, which
is more informative than `route_long_name` (not always populated in BODS
GTFS data). The V1 entries include both outbound trips (to Burton and
Wetmore) and inbound trips terminating at Derby Bus Station, since Bay 4
serves as both the origin and terminus for this route.

## Further reading

For timetable analysis, service pattern queries, and stop-level
departure boards, see the [tidytransit
vignettes](https://r-transit.github.io/tidytransit/articles/).
